Inequality in socioemotional skills: a cross-cohort comparison

### 2019-08-25 codes/R/paper/text/paper_jpube_3resultsconclusion.tex
text updates by Gabriella

Parameter estimates from our factor model are presented in \autoref{tab:measpars}. As discussed in the previous section, loadings and thresholds are constrained to have the same value across groups. Intercepts are normalised to zero, and error variances to one, for the reference group -- males in the BCS cohort. We use the estimates from this model to predict a score for each child in our sample along the latent externalising and internalising socio-emotional skill dimensions.\footnote{We use an empirical Bayes modal (EBM) approach to estimate the scores. The parameters are estimated using three sources of information. The first is the distribution of the latent variables $\bm{\theta}$, treated as random parameters with a prior $h(\bm{\theta}, \bm{\Omega})$, conditional on the parameters $\bm{\Omega}$. This prior is assumed to be multivariate normal. The second is the observed data $\bm{X}$, and the third is the estimated parameters $\hat{\bm{\Omega}}$. Data and prior are combined into the posterior distribution $w(\bm{\theta}| \bm{X}, \hat{\bm{\Omega}})$. For further details, see Chapter 7 in \citet{Skrondal2004}.} We plot the distribution of the scores in \autoref{fig:dens}. The unit of measurement is standard deviations of the distribution in the subsample of males in the BCS. Given our measurement invariance results in section \ref{sec:measinv}, we stress that the \emph{location} of these scores should not be directly compared across cohorts. However, the shape of the distribution can be given a cross-cohort interpretation.\footnote{The density of the scored factors can be contrasted with the distribution of sum scores in \autoref{fig:sumscores}. Using raw scores, instead, shows an increase in mass only at the top of the distribution.} It is immediately visible that there is more mass in the tails of the distribution in the 2000 than in the 1970 cohort.

@@

\autoref{fig:ineqsm} shows an even starker pattern when comparing children of mothers who smoked in pregnancy with non-smoking mothers. The fact that maternal smoking during pregnancy is a risk factor for offspring behavioural problems is well known in the medical literature \citep{Gaysina2013}; there is less evidence, however, on whether and to which extent these associations have changed across cohorts. The difference in child skills has increased, from less than .2 to around .4 of a standard deviation, again with the biggest increase experienced by the boys. There is also a significant increase in the gradient by paternal occupation based on social class (\autoref{fig:ineqfsc}), although this is less pronounced if compared to the one based on maternal characteristics. In particular, male children with no father figure living in their household have worse skills compared to children with blue collar fathers in the MCS cohort. Otherwise, skill differences in father's occupation are mostly constant across the two cohorts.\footnote{Figures \ref{fig:ineqips}, \ref{fig:ineqism}, and \ref{fig:ineqifsc} show inequality in the scale items underlying the factor scores used in this section. The increase in inequality across cohorts is still present, but less marked when looking at these single items. This shows the importance of the factor analysis step in aggregating items, explicitly modelling the measurement error, and testing and accounting for (loadings and thresholds) invariance across the two cohorts.} These patterns are in stark contrast with the findings of \citet{Reardon2016} for the US, who have found a narrowing of the readiness gaps from 1998 to 2010.

@@

Overall, the importance of maternal socioeconomic status (education and in particular employment) in determining socio-emotional skills has increased from the BCS to the MCS children. The `premium' in skills for children of better educated and employed mothers is significantly larger, for both boys and girls, internalising and externalising skills. At the same time, the penalty for having a blue-collar father, or not having a father figure at all in the household, has significantly declined across the two cohorts, especially for girls. Being born to an unmarried mother, and to a mother who smoked during pregnancy, is associated with a higher penalty for both dimensions of socio-emotional skills in the latter cohort, but only for males. This is consistent with recent evidence which shows that family disadvantage disproportionately impedes the pre-market development of boys, in terms of higher disciplinary problems, lower achievement scores, and fewer high-school completions \citep{Autor2016a}.\footnote{It is important to underscore that there has been a significant rise is cohabitation between 1975 and 2006. It is likely that unmarried mothers in the two cohorts have very different characteristics. The choice of this indicator is due to the absence of information on cohabitation in the birth survey for the BCS cohort.} Girls of non-white ethnicity, instead, have worse internalising and externalising skills in the MCS, a penalty not suffered by 5-year old non-white girls in the BCS. Firstborn boys and girls in the BCS have worse skills, but this difference disappears in the MCS. Lastly, we document an increase in the returns to birth weight, which is more pronounced for boys' internalising skills.

These changes in the relative importance of pregnancy factors and family background characteristics for child socio-emotional skills at age 5 need to be interpreted in the light of the significant changes in the prevalence of such characteristics across cohorts. As shown in \autoref{tab:sumstats}, the age of the mother at birth, proportion of mothers non-smoking in pregnancy, with post-compulsory education and in employment at the age 5 of the child has substantially increased; at the same time, the proportion of households with no father figure has increased, and so the proportion of women unmarried at birth is much higher in the 2000 than in the 1970 cohort. Also, the ethnic structure of the population has changed, with a higher proportion of non-white children in the MCS than in the BCS. In general, this has been a period of significant societal changes, with an almost continual rise in the proportion of women in employment, an older age at first birth and a rise in dual-earning parents families \citep{Roantree2018}. However, here we do not attempt to disentangle whether and to which extent the observed changes in inequality in socio-emotional skills across the two cohorts can be attributed to changes in returns (or penalties) to maternal characteristics (such as education and employment) or to compositional changes, as has been done for the analysis of wage inequality \citep{Blundell2007}.

@@

In this last section, we study the predictive power of socio-emotional skills for adolescent and adult outcomes. We contribute to a vast interdisciplinary literature by examining medium- and long-term impacts of skills measured at an earlier age than usually examined in previous studies, well before the start of formal education. We do so by regressing health and socioeconomic outcomes measured in adolescence and adulthood on the socio-emotional skills scores at age five obtained by our factor model, controlling for the harmonised family background variables at birth and age five (see \autoref{tab:harmvar}).\footnote{In \autoref{tab:bcs_adultapp}, we show that the conclusions in this section are not sensitive to the factor scoring methodology used. 'Raw' score, obtained by simple summing of the item categories in the 11-item subscale have basically equal predictive power to factor scores.} We present results with and without controlling for cognitive skills. As detailed in \autoref{sec:data}, the available cognitive measures are not comparable across cohorts. Still, we control for a factor score that summarises all information on cognitive skills that is available in each cohort, regardless of their comparability.

Socio-emotional skills at five years of age are predictive of adolescent health behaviour and outcomes in both cohorts.\footnote{Unfortunately the strength of the association cannot be directly compared, since the outcomes are measured at different ages: 16 and 14 years for BCS and MCS, respectively.} \autoref{tab:adol} examines adolescent smoking and BMI for both cohorts; \autoref{tab:bcs_adultapp} reports the results for the same outcomes in adulthood (at age 42), for the BCS only. Externalising skills are negatively correlated to subsequent smoking and BMI in both cohorts, for both genders. Recall that a child with high externalising skills exhibits less restless and hyperactive behaviours, and has less anti-social conduct. Our findings are consistent with the body of evidence reviewed in section \ref{sec:lit}, which shows that better socio-emotional skills (measured using different scales and at various points during childhood and adolescence) are negatively associated with smoking. At the same time, internalising skills are positively correlated with smoking (only in the 1970 cohort) and BMI (only for girls), although less strongly than externalising skills. This apparently counterintuitive result makes sense in light of the items in our internalising scale shown in \autoref{tab:scalecomp}. A child with better internalising skills is less solitary, neurotic, and worried. From this perspective, she is likely more sociable and subject to peer influence in health behaviours. This is consistent with the evidence in \citet{Goodman2015}, who find a positive association between child emotional health (measured with items for the internalizing behaviour subscale of the Rutter scale at age 10 in the BCS) and smoking at age 42. Furthermore, in recent work \citet{Hsieh2018} have shown personality to be a key mechanism through which peers affect smoking behaviour.

Conditional on socio-emotional skills, cognition has limited predictive power for these behaviours, and only for girls.\footnote{We do not observe significant associations between early socio-emotional skills and other risky behaviours like drug-taking and alcohol consumption. One possible reason might be the relatively young age at which these skills are measured. Results are available upon request.} This is in line with the evidence in \citet{Conti2010}, who show that not accounting for non-cognitive traits (a self-regulation factor measured at age 10) overestimates the importance of cognition for predicting health and health behaviours, using data from the British cohort study. \citet{Conti2013a} use rich data on child personality and socio-emotional traits collected at ages 7, 11 and 16 in the 1958 British birth cohort,\footnote{They use the Rutter scale and the Bristol Social Adjustment Guide.} and show that these traits rival the importance of cognition in explaining the education gradient in health behaviours (including smoking and BMI). We show that child socio-emotional skills have greater predictive power for health outcomes and behaviours even when measured at an earlier age.

Cohort members from the British Cohort Study are now well into their adulthood. For this cohort, we can examine the association between socio-emotional skills at age five and adult education and labour market outcomes. The structure of \autoref{tab:bcs_adult} is similar to \autoref{tab:adol}, but it considers educational achievement, employment, and earnings (conditional on being in paid employment) for the BCS cohort members. For these outcomes, the predictive power of cognitive skills outweighs that of socio-emotional skills, whose predictive power diminishes over time (between the ages 34 and 42), and is driven to insignificance after controlling for cognition. This is consistent with the evidence in  \citet{Conti2011}, who show that cognitive endowments at age 10 are more predictive (than socio-emotional and health ones) for employment and wage outcomes in the BCS. Again, we show that the greater predictive power of cognition for socioeconomic outcomes holds even when considering earlier-life measures of child development. Similarly, \citet{Goodman2015} use data from the same cohort, but exploit a richer set of measures from the age 10 survey. They show that social and emotional skills measured at age 10 are strongly related to a wide range of adult outcomes.

@@

Third, after having computed comparable scores for both externalising and internalising skills, and for both boys and girls, we have compared how inequality in these skills has changed across the 1970 and the 2000 cohort. We have documented for the first time that inequality in these early skills has increased across cohorts, especially for boys. The cross-cohort increase in the gap is more pronounced at the bottom of the distribution (50-10 percentiles). We have also documented changes in conditional skills gaps across cohorts. In particular, the difference in the socio-emotional skills of their children between mothers of higher and lower socio-economic status (education and employment) has increased. The increase in cross-cohort inequality is even starker when comparing children born to mothers who smoked during pregnancy. In both cases, the increase in inequality is particularly pronounced for boys. On the other hand, the skills penalty arising from the lack of a father figure in the household has substantially declined, especially for girls.

>>>>

Parameter estimates from our factor model are presented in \autoref{tab:measpars}. As discussed in the previous section, loadings and thresholds are constrained to have the same value across groups. Intercepts are normalised to zero, and error variances to one, for the reference group -- males in the BCS cohort. We use the estimates from this model to predict a score for each child in our sample along the latent externalising and internalising socio-emotional skill dimensions.\footnote{We use an empirical Bayes modal (EBM) approach to estimate the scores. The parameters are estimated using three sources of information. The first is the distribution of the latent variables $\bm{\theta}$, treated as random parameters with a prior $h(\bm{\theta}, \bm{\Omega})$, conditional on the parameters $\bm{\Omega}$. This prior is assumed to be multivariate normal. The second is the observed data $\bm{X}$, and the third is the estimated parameters $\hat{\bm{\Omega}}$. Data and prior are combined into the posterior distribution $w(\bm{\theta}| \bm{X}, \hat{\bm{\Omega}})$. For further details, see Chapter 7 in \citet{Skrondal2004}.} We plot the distribution of the scores in \autoref{fig:dens}. The unit of measurement is standard deviations of the distribution in the subsample of males in the BCS. Given our measurement invariance results in section \ref{sec:measinv}, we stress that the \emph{location} of these scores should not be directly compared across cohorts. However, the shape of the distribution can be given a cross-cohort interpretation. This result is in sharp contrast with what shown by the simple distribution of sum scores in \ref{fig:sumscores}: using raw scores shows an increase in mass only at the top of the distribution. It is immediately visible that there is more mass in the tails of the distribution in the 2000 than in the 1970 cohort.

@@

\autoref{fig:ineqsm} shows an even starker pattern when comparing children of mothers who smoked in pregnancy with non-smoking mothers. The fact that maternal smoking during pregnancy is a risk factor for offspring behavioural problems is well known in the medical literature \citep{Gaysina2013}; there is less evidence, however, on whether and to which extent these associations have changed across cohorts. The difference in child skills has increased, from less than .2 to around .4 of a standard deviation, again with the biggest increase experienced by the boys. There is also a significant increase in the gradient by paternal occupation based on social class (\autoref{fig:ineqfsc}), although this is less pronounced if compared to the one based on maternal characteristics. In particular, male children with no father figure living in their household have worse skills (both internalising and externalising) compared to children with blue collar fathers in the MCS cohort. Otherwise, skill differences in father's occupation are mostly constant across the two cohorts.\footnote{Figures \ref{fig:ineqips}, \ref{fig:ineqism}, and \ref{fig:ineqifsc} show inequality in the scale items underlying the factor scores used in this section. The increase in inequality across cohorts is still present, but less marked when looking at these single items. This shows the importance of the factor analysis step in aggregating items, explicitly modelling the measurement error, and testing and accounting for (loadings and thresholds) invariance across the two cohorts.} These patterns are in stark contrast with the findings of \citet{Reardon2016} for the US, who have found a narrowing of the readiness gaps from 1998 to 2010 (however, they have not tested for measurement invariance).

@@

Overall, the importance of maternal socioeconomic status (education and in particular employment) in determining socio-emotional skills has increased from the BCS to the MCS children. The `premium' in skills for children of better educated and employed mothers is significantly larger, for both boys and girls, internalising and externalising skills. At the same time, the penalty for having a blue-collar father, or not having a father figure at all in the household, has significantly declined across the two cohorts, especially for girls. Being born to an unmarried mother, and to a mother who smoked during pregnancy, is associated with a higher penalty for both dimensions of socio-emotional skills in the latter cohort.\footnote{It is important to underscore that there has been a significant rise is cohabitation between 1975 and 2006. It is likely that unmarried mothers in the two cohorts have very different characteristics. The choice of this indicator is due to the absence of information on cohabitation in the birth survey for the BCS cohort.} Children of non-white ethnicity have worse internalising and externalising skills in the MCS, a penalty almost absent in the BCS (where the prevalence of non-white children was much lower). Firstborn boys and girls in the BCS have worse skills, but this difference disappears in the MCS. Lastly, we document an increase in the returns to birth weight, which is more pronounced for boys.

These changes in the relative importance of pregnancy factors and family background characteristics for child socio-emotional skills at age 5 need to be interpreted in the light of the significant changes in the prevalence of such characteristics across cohorts. As shown in \autoref{tab:sumstats}, the age of the mother at birth, and the proportion of mothers non-smoking in pregnancy, with post-compulsory education and in employment at the age 5 of the child has substantially increased; at the same time, the proportion of households with no father figure has increased, and so the proportion of women unmarried at birth is much higher in the 2000 than in the 1970 cohort. Also, as noted, the ethnic structure of the population has changed, with a higher proportion of non-white children in the MCS than in the BCS. In general, this has been a period of significant societal changes, with an almost continual rise in the proportion of women in employment, an older age at first birth and a rise in dual-earning parents families \citep{Roantree2018}. However, here we do not attempt to disentangle whether and to which extent the observed changes in inequality in socio-emotional skills across the two cohorts can be attributed to changes in returns (or penalties) to maternal characteristics (such as education and employment) or to compositional changes, as has been done for the analysis of wage inequality \citep{Blundell2007}.

@@

In this last section, we study the predictive power of socio-emotional skills for adolescent and adult outcomes. We contribute to a vast interdisciplinary literature by examining medium- and long-term impacts of skills measured at an earlier age than usually examined in previous studies, i.e. well before the start of formal education. We do so by regressing health and socioeconomic outcomes measured in adolescence and adulthood on the socio-emotional skills scores at age five obtained by our factor model, controlling for the harmonised family background variables at birth and age five (see \autoref{tab:harmvar}).\footnote{In tables \ref{tab:adol_sum} and \ref{tab:bcs_adultapp}, we show that the conclusions in this section are not sensitive to the factor scoring methodology used. 'Raw' scores, obtained by a simple unweighted average of the item categories in the 11-item subscale have basically equal predictive power to factor scores.} We present results with and without controlling for cognitive skills. As detailed in \autoref{sec:data}, the available cognitive measures are not comparable across cohorts. Still, we control for a factor score that summarises all information on cognitive skills that is available in each cohort, regardless of their comparability.

Socio-emotional skills at five years of age are predictive of adolescent health behaviour and outcomes in both cohorts.\footnote{Unfortunately the strength of the association cannot be directly compared, since the outcomes are measured at different ages: 16 and 14 years for BCS and MCS, respectively.} \autoref{tab:adol} examines adolescent smoking and BMI for both cohorts; \autoref{tab:bcs_adultapp} reports the results for the same outcomes in adulthood (at age 42), for the BCS only. Externalising skills are negatively correlated to subsequent smoking and BMI in both cohorts, for both genders. Recall that a child with high externalising skills exhibits less restless and hyperactive behaviours, and has less anti-social conduct. Our findings are consistent with the body of evidence reviewed in section \ref{sec:lit}, which shows that better socio-emotional skills (measured using different scales and at various points during childhood and adolescence) are negatively associated with smoking. At the same time, internalising skills are positively correlated with smoking (only in the 1970 cohort) and BMI (only for girls), although less strongly than externalising skills. This apparently counterintuitive result makes sense in light of the items in our internalising scale shown in \autoref{tab:scalecomp}. A child with better internalising skills is less solitary, neurotic, and worried. From this perspective, he/she is likely more sociable and subject to peer influence in health behaviours. This is consistent with the evidence in \citet{Goodman2015}, who find a positive association between child emotional health (measured with items for the internalizing behaviour subscale of the Rutter scale at age 10 in the BCS) and smoking at age 42. Furthermore, in recent work \citet{Hsieh2018} have shown personality to be a key mechanism through which peers affect smoking behaviour.

Conditional on socio-emotional skills, cognition has limited predictive power for these behaviours, and only for girls.\footnote{We do not observe significant associations between early socio-emotional skills and other risky behaviours like drug-taking and alcohol consumption. One possible reason might be the relatively young age at which these skills are measured. Results are available upon request.} This is in line with the evidence in \citet{Conti2010}, who show that not accounting for non-cognitive traits (in their paper, a self-regulation factor measured at age 10) overestimates the importance of cognition for predicting health and health behaviours, using data from the British cohort study. Along the same lines, \citet{Conti2013a} use rich data on child personality and socio-emotional traits collected at ages 7, 11 and 16 in the 1958 British birth cohort,\footnote{They use the Rutter scale and the Bristol Social Adjustment Guide.} and show that these traits rival the importance of cognition in explaining the education gradient in health behaviours (including smoking and BMI). We show that child socio-emotional skills have greater predictive power for health outcomes and behaviours even when measured at an earlier age.

Cohort members from the British Cohort Study are now well into their adulthood. For this cohort, we can examine the association between socio-emotional skills at age five and adult education and labour market outcomes. The structure of \autoref{tab:bcs_adult} is similar to \autoref{tab:adol}, but it considers educational achievement, employment, and earnings (conditional on being in paid employment) for the BCS cohort members. For these outcomes, the predictive power of cognitive skills outweighs that of socio-emotional skills, which are only predictive of educational attainment, and whose predictive power for males is driven to insignificance after controlling for cognition. This is consistent with the evidence in  \citet{Conti2011}, who show that cognitive endowments at age 10 are more predictive (than socio-emotional and health ones) for employment and wage outcomes in the BCS. Again, we show that the greater predictive power of cognition for socioeconomic outcomes holds even when considering earlier-life measures of child development.

@@

Third, after having computed comparable scores for both externalising and internalising skills, and for both boys and girls, we have compared how inequality in these skills has changed across the 1970 and the 2000 cohort. We have documented for the first time that inequality in these early skills has increased across cohorts, especially for boys. The cross-cohort increase in the gap is more pronounced at the bottom of the distribution (50-10 percentiles). We have also documented changes in conditional skills gaps across cohorts. In particular, the difference in the socio-emotional skills of their children between mothers of higher and lower socio-economic status (education and employment) has increased. The increase in cross-cohort inequality is even starker when comparing children born to mothers who smoked during pregnancy. On the other hand, the skills penalty arising from the lack of a father figure in the household has substantially declined.

### 2019-08-25 codes/R/paper/text/paper_jpube_4appendix.tex
text updates by Gabriella

In the BCS data, maternal reports on child socioemotional skills are measured using the Rutter A Scale \citep{Rutter1970} -- see Panel A of \autoref{tab:scales}. The Rutter items are rated on three levels: `Does not apply', `Somewhat applies', `Certainly applies'. Since they are all behaviours indicating lower skills, we encode all of them in reverse, i.e. `Certainly applies' = 0, `Somewhat applies' = 1, `Does not apply' = 2. We augment the 19-item Rutter Scale with three additional parent-reported questions from the parental questionnaire, items A, B, and D. These are rated on 4 levels: `Never in the last 12 months', `less than once a month', `at least once a month', `at least once a week'. we recode these into binary indicators, with `Never' and `Less than once a month' to 1 and zero otherwise. To increase comparability between the two scales, we merge together two pairs of items: 4 and 19 (to mirror SDQ item 12 ``Often fights with other children or bullies them"), and A and B (to mirror SDQ item 3 ``Often complains of head-aches, stomach-ache or sickness"). We assign the lowest category among the two original items to the newly obtained item. We also recode items 5 and 14 to binary instead of three categories. These items are recorded with a positive phrasing in SDQ, so a 3-category split would be harder to compare.

In MCS, we use the 25-item strengths and difficulties questionnaire \citep{Goodman1997} -- see Panel B of \autoref{tab:scales}. All items are recorded on a 4-point scale: `Not true', `Somewhat true', `Certainly true', `Can't say'. We set the latter option to missing and recode the rest in ascending order of skill as for the BCS items, i.e. `Certainly true' = 0, `Somewhat true' = 1, `Not true' = 2. For comparability with the BCS Rutter scale, we dichotomise items 3 and 5 to make them comparable with , and dichotomise and invert items 7, and 14.

@@

In fact, exploratory factor analysis (on both on the full Rutter and SDQ scales and on the 11-item subscale) shows that these items, numbered 5 and 11, perform poorly and exhibit relatively low factor loadings. As a robustness check, we replicate the main results of the paper by excluding them from the subscale.

>>>>

In the BCS data, maternal reports on child socioemotional skills are measured using the Rutter A Scale \citep{Rutter1970} -- see Panel A of \autoref{tab:scales}. The Rutter items are rated on three levels: `Does not apply', `Somewhat applies', `Certainly applies'. Since they are all behaviours indicating lower skills, we recode all of them in reverse, i.e. `Certainly applies' = 0, `Somewhat applies' = 1, `Does not apply' = 2. We augment the 19-item Rutter Scale with three additional parent-reported questions from the parental questionnaire, items A, B, and D. These are rated on 4 levels: `Never in the last 12 months', `less than once a month', `at least once a month', `at least once a week'. we recode these into binary indicators, with `Never' and `Less than once a month' to 1 and zero otherwise. To increase comparability between the two scales, we merge together two pairs of items: 4 and 19 (to mirror SDQ item 12 ``Often fights with other children or bullies them"), and A and B (to mirror SDQ item 3 ``Often complains of head-aches, stomach-ache or sickness"). We assign the lowest category among the two original items to the newly obtained item. We also recode items 5 and 14 to binary instead of three categories. These items are recorded with a positive phrasing in SDQ, so a 3-category split would be harder to compare.

In MCS, we use the 25-item Strengths and Difficulties Questionnaire \citep{Goodman1997} -- see Panel B of \autoref{tab:scales}. All items are recorded on a 4-point scale: `Not true', `Somewhat true', `Certainly true', `Can't say'. We set the latter option to missing and recode the rest in ascending order of skill as for the BCS items, i.e. `Certainly true' = 0, `Somewhat true' = 1, `Not true' = 2. For comparability with the BCS Rutter scale, we dichotomise items 3 and 5, and dichotomise and invert items 7, and 14.

@@

In fact, exploratory factor analysis (on both the full Rutter and SDQ scales and on the 11-item subscale) shows that these items, numbered 5 and 11, perform poorly and exhibit relatively low factor loadings. As a robustness check, we replicate the main results of the paper by excluding them from the subscale.

### 2019-08-25 codes/R/paper/text/paper_jpube_5apptables.tex
text updates by Gabriella

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{landscape}
\begin{table}[ht!]
\caption{Description of harmonised variables}\label{tab:harmvar}
\centering\setstretch{1.2} \scriptsize
\begin{tabular}{L{.18\textwidth}C{.1\textwidth}L{.3\textwidth}L{.75\textwidth}}
\toprule
Variable Group & Age & Variable & Note \\
\midrule
Maternal education & 5 & Post-compulsory schooling$^d$ & Whether mother continued schooling past the compulsory age, based on her year of birth. School leaving age in England was changed from 14 to 15 in 1947 and from 15 to 16 in 1972. \\ \\[-1.5em]
Maternal employment & 5 & Employed$^d$ & Includes full time and part time \\ \\[-1.5em]
Father occupation & 5 & White collar (I-IIINM)$^d$ \newline Blue collar (IIIM-V-other)$^d$ \newline No father figure$^d$ & Based on father's Registrar General Social Class classification of occupations. White collar includes I (Professional), II (Managerial/technical), IIINM (Skilled non-manual). Blue collar includes IIIM (Skilled manual), IV (Partly skilled), V (Unskilled), Other, Unemployed, and Armed forces. No father figure is a dummy for children whose father does not live in the same household. Father's social class was recorded using the SOC2000 classification in MCS. We use the derivation matrices kindly provided by David Pevalin at ISER (available at https://www.iser.essex.ac.uk/archives/nssec/derivations-of-social-class) to map SOC2000 into Registrar General Social Class. \\ \\[-1.5em]
Maternal background & 0/5 & Mother's age at birth \newline Mother's height (cm) \newline Mother unmarried at birth$^d$ \newline Child nonwhite ethnicity$^d$  \newline Number of children in HH & All variables are self-reported by the mother at birth, except for number of children in household (at five years old). Unmarried is only based on marital status, and includes cohabitation. \\ \\[-1.5em]
Pregnancy & 0 & Child is firstborn$^d$ \newline Number of previous stillbirths \newline Mother smoked in pregnancy$^d$ \newline Preterm birth (under 37 weeks gestation)$^d$ \newline (log) birth weight (kg) & Parity, stillbirths, and smoking are self-reported by the mother. Gestational length and birth weight are from hospital records. \\ \\[-1.5em]
Cognitive skills & 5 & & Based on test batteries administered to the cohort member at five. Three tests are used for BCS children: Copy Designs child is asked to copy simple designs adjacently), Human Figure Drawing (child draws an entire human figure), English Picture Vocabulary Test (child identifies the picture referring to a word among four pictures). Three \emph{different} tests are used in the MCS: BAS Naming Vocabulary (child is shown a series of pictures and asked to name it), BAS Picture Similarity (child is shown a row of 4 pictures on a page and places a card with a fifth picture under the one most similar to it), BAS Pattern Construction (child constructs a design by putting together flat squares or solid cubes with patterns on each side).

 \\ \\[-1.5em]
Adolescent outcomes & 16 (BCS) \newline 14 (MCS) & Child tried smoking$^d$ \newline Body Mass Index (BMI) & Smoking is self reported by the child. Height and weight are taken as part of a medical examination. \\ \\[-1.5em]
Adult outcomes (BCS only) & 34 \newline 42 \newline 34, 42 & Higher education$^d$ \newline Employed$^d$ \newline (log) gross weekly pay & Higher education is defined on having a university degree or its vocational equivalent. It corresponds to level 4 or 5 in the National Vocational Qualification (NVQ) equivalence. Employed is a dummy for being in paid employment or self-employment, either full or part time. Gross weekly pay is weekly pre-tax pay from the respondent's main activity, conditional on being a paid employee. \\
\bottomrule
\end{tabular}
\justify {\scriptsize \emph{Notes}: Variables denoted by $^d$ are binary or categorical.
\par}
\end{table}
\end{landscape}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{sidewaystable}[ht!]
\caption{Description of harmonised variables}\label{tab:harmvar}
\centering\setstretch{1.2} \scriptsize
\begin{tabular}{L{.18\textwidth}C{.1\textwidth}L{.3\textwidth}L{.75\textwidth}}
\toprule
Variable Group & Age & Variable & Note \\
\midrule
Maternal education & 5 & Post-compulsory schooling$^d$ & Whether mother continued schooling past the compulsory age, based on her year of birth. School leaving age in England was changed from 14 to 15 in 1947 and from 15 to 16 in 1972. \\ \\[-1.5em]
Maternal employment & 5 & Employed$^d$ & Includes full time and part time \\ \\[-1.5em]
Father occupation & 5 & White collar (I-IIINM)$^d$ \newline Blue collar (IIIM-V-other)$^d$ \newline No father figure$^d$ & Based on father's Registrar General Social Class classification of occupations. White collar includes I (Professional), II (Managerial/technical), IIINM (Skilled non-manual). Blue collar includes IIIM (Skilled manual), IV (Partly skilled), V (Unskilled), Other, Unemployed, and Armed forces. No father figure is a dummy for children whose father does not live in the same household. Father's social class was recorded using the SOC2000 classification in MCS. We use the derivation matrices kindly provided by David Pevalin at ISER (available at https://www.iser.essex.ac.uk/archives/nssec/derivations-of-social-class) to map SOC2000 into Registrar General Social Class. \\ \\[-1.5em]
Maternal background & 0/5 & Mother's age at birth \newline Mother's height (cm) \newline Mother unmarried at birth$^d$ \newline Child nonwhite ethnicity$^d$  \newline Number of children in HH & All variables are self-reported by the mother at birth, except for number of children in household (at five years old). Unmarried is only based on marital status, and includes cohabitation. \\ \\[-1.5em]
Pregnancy & 0 & Child is firstborn$^d$ \newline Number of previous stillbirths \newline Mother smoked in pregnancy$^d$ \newline Preterm birth (under 37 weeks gestation)$^d$ \newline (log) birth weight (kg) & Parity, stillbirths, and smoking are self-reported by the mother. Gestational length and birth weight are from hospital records. \\ \\[-1.5em]
Cognitive skills & 5 & & Based on test batteries administered to the cohort member at five. Three tests are used for BCS children: Copy Designs (child is asked to copy simple designs adjacently), Human Figure Drawing (child draws an entire human figure), English Picture Vocabulary Test (child identifies the picture referring to a word among four pictures). Three tests are used in the MCS: BAS Naming Vocabulary (child is shown a series of pictures and asked to name it), BAS Picture Similarity (child is shown a row of 4 pictures on a page and places a card with a fifth picture under the one most similar to it), BAS Pattern Construction (child constructs a design by putting together flat squares or solid cubes with patterns on each side).
 \\ \\[-1.5em]
Adolescent outcomes & 16 (BCS) \newline 14 (MCS) & Child tried smoking$^d$ \newline Body Mass Index (BMI) & Smoking is self reported by the child. Height and weight are taken as part of a medical examination. \\ \\[-1.5em]
Adult outcomes (BCS only) & 34 \newline 42 \newline 34, 42 & Higher education$^d$ \newline Employed$^d$ \newline (log) gross weekly pay & Higher education is defined as having a university degree or its vocational equivalent. It corresponds to level 4 or 5 in the National Vocational Qualification (NVQ) equivalence. Employed is a dummy for being in paid employment or self-employment, either full or part time. Gross weekly pay is weekly pre-tax pay from the respondent's main activity, conditional on being a paid employee. \\
\bottomrule
\end{tabular}
\justify {\scriptsize \emph{Notes}: Variables denoted by $^d$ are binary or categorical.
\par}
\end{sidewaystable}

@@
